---
title: "Dynamic stuff with R Markdown, HTML"
author: "CCO, JA"
date: "October 12, 2015"
output:
  html_document:
    highlight: haddock
    theme: spacelab
    toc: yes
  pdf_document:
    highlight: haddock
    toc: yes
---

```{r setup, echo = FALSE, eval = TRUE, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(DT)
library(knitr)

```

There are some cool stuff you can do with `Rmarkdown` when you `knit` it to .html. Among the most cool stuff that you can do are dynamic outputs (tables and figures).

## Dynamic tables

For this example we're gonna use our classic [Alaska](http://www.seaaroundus.org/?gclid=Cj0KEQjw8tbHBRC6rLS024qYjtEBEiQA7wIDeUk0hWAQ_MkdhsIBczR7jaN3qC4ndTuExVUNThk0g2kaArm88P8HAQ) fisheries data from Sea Around Us. There are probably many packages that can be used to show dynamic tables, I know [datatable](https://rstudio.github.io/DT/) so will show some examples:

### Normal Table

```{r Table_example, echo=F, eval=F}

Alaska_Data <- read.csv("data/Alaska.csv")

datatable(Alaska_Data[1:5])

```

### Other Options

```{r Table_example_II, echo=T, eval=T}

Alaska_Data <- read.csv("data/Alaska.csv")

datatable(Alaska_Data[1:5],
          caption = 'Table 1: This is a simple caption for the table.',
          filter = 'top', options = list(
  pageLength = 5, autoWidth = TRUE
)
)

```



### Dynamic Maps

Remember our map example of last session? Well, you can use [leaflet](https://rstudio.github.io/leaflet/markers.html) to show dynamic maps in R-markdown! Say whaaaaaaaaaat?

```{r leafletMap, echo=F,eval=T, message=F}

library(leaflet)
library(htmltools)

data <- quakes

FEDECOOP <- paste(sep = "<br/>",
                    "<b><a href='http://www.fedecoop.com.mx'>MetaID 342</a></b>",
                    "Lobster Stock Assesment and Catch from FEDECOOP since 1970",
                    "Private Dataset"
  )

leaflet(data) %>%
      addTiles(
        urlTemplate = "//{s}.tiles.mapbox.com/v3/jcheng.map-5ebohr46/{z}/{x}/{y}.png",
        attribution = 'Maps by <a href="http://www.mapbox.com/">Mapbox</a>'
      ) %>%
      #Initial view #
      setView(lng = -102.5528, 
              lat = 23.6345,
              zoom = 5) %>% 
      # Data examples as markers ####
    addCircleMarkers(lng = -109.4725,
               lat = 24.6356, 
               popup = "7654")%>%
      addMarkers(lng = -95.8084852,
                 lat = 20.5764432,
                 popup = "Where I lost my Dive watch"
      )%>%
      addMarkers(lng = -107.8084852,
                 lat = 15.5764432,
                 popup = "Collection point"
      )%>%
      # Data examples as rectangles ####
    addRectangles(
      lng1=-107.607877, lat1=18.459820,
      lng2=-110.557877, lat2=17.431188,
      color="red",
      fillColor = "red",
      fillOpacity = 0,
      popup = "The We all love fish MPA; Since 1980"
    ) %>% 
      addRectangles(
        lng1=-115.507877, lat1=28.400000,
        lng2=-115.557877, lat2=28.421188,
        color="green",
        fillColor = "green",
        fillOpacity = 0.2,
        popup = FEDECOOP
      ) %>% 
  addTiles() %>%
  addMarkers(
  clusterOptions = markerClusterOptions()
)
  
```