---
always_allow_html: yes
layout: page
title: 'Introduction to Statistics in R'
visible: false
tags:
  - r
  - intermediate
  - statistics
---

* Author: Lindsay Coome
* Lesson Topic: Introduction to Statistics in R
* Lesson video stream: <https://www.youtube.com/watch?v=A7AD40fpLK8>


```{r}
# Have the following packages installed and loaded:
# install.packages("car")
# install.packages("psych")
# install.packages("multcomp")
library(car)
library(psych)
library(multcomp)
```

## Descriptive statistics
### Frequencies with table():
The table() function displays counts of identical observations for either a single data vector or a dataframe

```{r}
#table
table(mtcars$gear)
#cross tabulate
#note: first variable is set as the rows
table(mtcars$gear, mtcars$carb)
```


### Mean:
Without the na.rm=T argument, mean() will return a null value if there is any missing data in the object

```{r}
mean(Wong$viq, na.rm=T)
```

### Median:

```{r}
median(Wong$viq, na.rm=T)
```

### Variance and SD:

```{r}
var(Wong$viq, na.rm=T)
sd(Wong$viq, na.rm=T)
```

### Range: 

```{r}
range(Wong$viq, na.rm=T)
```

### describe() from the psych package:
Provides item name, item number, nvalid, mean, sd, median, mad (median absolute deviation from the median), min, max, skew, kurtosis, se

```{r}
describe(Wong$age)
```


### describeBy() from the psych package:
A simple way of generating summary statistics by grouping variable

```{r}
describeBy(Wong$age, Wong$sex)
```

## Inferential statistics
### One-sample t test:

```{r}
t.test(Wong$age, mu=30)
```

### Two sample t-test:

```{r}
#one way to compare groups:
#t.test( object1, object1)
#or use sex as the grouping variable:
t.test(Wong$age~Wong$sex)
```

### Paired samples t-test:

```{r}
#t.test(object1, object1, paired=TRUE)
```

### Correlations:
cor.test gives the correlation and significance test on the correlation

```{r}
cor.test(iris$Petal.Length, iris$Petal.Width)
```

### General linear model (i.e., with a normally-distributed dependent variable):
Dependent variables are predicted by a tilde “~” 
The formula to regress “y” on “x” is y ~ x

```{r}
result <- lm(viq~duration, data=Wong)
summary(result)

#multiple predictors with interaction terms
result <- lm(viq~duration*days*sex, data=Wong)
summary(result)

#use + for main effect of predictor only
result <- lm(viq~duration*days + sex, data=Wong)
summary(result)
```

### Analysis of Variance:
ANOVA is simply a different way of evaluating explained variance in linear modelling
You must always wrap the anova() function around a lm() function

```{r}
anova(lm(mpg ~ am*gear, data=mtcars))
```

However, R uses type II sums of squares by default
Anova() from the car package can give you “Type III Sums of Squares” 
This matters when you have more than one predictor (e.g. treatment x sex)
Asking for type III sums of squares will match what you get from SPSS or SAS

```{r}
result <- lm(mpg ~ am*gear, data=mtcars)
Anova(result, type=3)
```

### Post hoc tests:
R comes with a default pairwise t-test (pairwise.t.test())
multcomp offers more posthoc options than base R

```{r}
result <- lm(Petal.Length ~ Species, data=iris)
postHocs<-glht(result, linfct = mcp(Species = "Tukey"))
#summary function gives results of multiple comparisons
summary(postHocs)
#you can also ask for confidence intervals
confint(postHocs)
```

### Logistic regression:
Normally-distributed dependent variable assumption violated
So, you must use the glm() function rather than lm()

```{r}
summary(glm(Wong$sex ~ Wong$piq*Wong$viq, family=binomial))
```

### Scales and reliability:
Learn how to sum across rows (useful for getting a summary score from a scale)
Can get Cronbach's Alpha from the ```psych``` package

```{r}
#Subset iris data so only numerical columns remain
newiris <- iris[c(1:4)]
#Calculate row means
rowMeans(newiris, na.rm=TRUE)
#Use the psych package for Cronbach's Alpha
alpha(newiris)
```









