---
title: "Intro_R_Oct8_2021"
author: "Tara"
date: "10/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Welcome

 - **Authors**: Tara Henechowicz adapted from plans by Margot Lautens and Luke Johnston
 - **Research field**: Neuroscience (Music and Health)
 - **Lesson topic**: Intro to R, some data wrangling, statistics, and some visualization
 - **Lesson content URL**: <https://github.com/UofTCoders/studyGroup/tree/gh-pages/lessons/r/intro>

#Why should we use R? 
- reproducible science 
- wonderful packages to help you for working with your data (even datasets that are very big)!
- packages specific to disciplines: biology, genetics, even neuroimaging
- can customize and create reports with RMarkdown. 

#What can we do with R? 
- Organize: Tidy data
- Analyze: statistics, machine learning, field-specific techniques
- Visualize

#Starting with R - the command line

##Command Line

The command line executes code like a calculator but does not keep a record of your work. 
Try entering 2+2 in the commandline, what is the output? 

Now lets save 4 as a variable a. The way we do this in R is typing `a <- 4` in the command line. Here the `<-` is acting as an = sign. 

In summary, you can execute code in the command line, but you will need to scroll back to see the previous code you executed. To reproduce your code, you will need to create a script. 

##Scripts
The type of script we are working in is called and .Rmd file it is an R Markdown script. THis script is helpful for building reports which can later be rendered as a pdf or html file. 

Below is a `code chunk`, where you keep a record of your code and you can execute that code. Anything outside of the code can be used as text to supplement your report or write notes about your process. You can open a code chunk using the `Insert` button or using the short cut for your device. 

Try running the code chunk below by pressing the green play button. 
```{r}
#this is a comment, this will not run as code. 
#Assign the value of 4 to variable "a"
a<-4
#return the value of a
print(a)
#Ask r if 2+2 is equal to "a"
2+2==a
```

##Other windows in R
- Environment: all the variables we have saved, our functions, and libraries we have loaded
- History
- Files
- Help 

#Starting your script
1. Set our "working directory", this means we are telling R where to take files from and where to write them. 
2. Load in the packages we need to use for our project. 
If we don't have the package installed already, you will use the `install.packages` function.
Whether you have the package or not, you will load the package using the `library` function.
```{r}
#Load in the tidyverse and Using Rlibrary
library(tidyverse)
library(UsingR)
```

3. Load in our data.
We will be using a dataset that is already in R, so lets take a look at `babies` and try to store as variable `df` in our environment.
```{r}
#call the babies dataset
babies
#we still need to store this dataset in our R environment
df <- babies
```

Now if we call df, we can see we have the babies dataset 
```{r}
#viewing df
view(df)
#OR
df
```

Lets use the function write.csv to take that data and save it as a .csv file in our current directory

```{r}
write.csv(babies, "babies.csv")
```

A few ways to lad data are to:
- manually use the import dataset button in Rstudio (and copy the code from the output)
- use `read_csv()` function from readr package
```{r}
#copy the code generated by the import dataset button
library(readr)
babies2 <- read_csv("babies.csv")
View(babies)
```

#Viewing the data and its structure

1. Instead of viewing the whole dataset we can use the `head` function to view the first 6 rows of the data. 
2. If you want to view more rows of data, use the argument n= to change the size of the table returned. 
```{r}
#view the whole dataset
view(df)
#use the head function to view the first 6 rows of the data
head(df)
#Look at the first 10 rows of the data
head(df, n=10)
```

3. Another important function is `str` it returns the structure of the object. For a `data.frame` that we have here, it will also give you the column names and the structure of those columns. 
`dim` is also a useful function that will retrun the dimesnions of your data.frame.
`names()` shows the column names.
`glimpse()` is a `dplyr` only function that ressembles `str` but unlike `str`
it allows you to see all your columns.
`row.names()` or `rownames()` will return a vector of the rownames of the dataset. However, only `rownames()`can be changed, we will return to the uses of this function later. 

```{r}
#use the str() function on df
str(df)
#use the dim() function to return the size of df
dim(df)
#use names() to return the column names
names(df)
#use row.names() or rownames() to return a vector of the rownames
row.names(df)
```

#Summary statistics and exploring patterns

1. `summary()` shows a quick description of the summary statistics (means, 
median, frequency)for each of your columns.  

This function is especially useful for finding issues in the data coding and in the data structure.
For example, look at the variable sex - is there is something strange here?
Are there any other variables that might have issues?
These are all issues we can fix when we tidy the data, we will return to data tidying in the next section. 
```{r}
#give a summary of df
summary(df)
```

2. Plotting to explore the data

Before we plot any of the columns, we need to know how to select columns from the dataset. 
To call a column of the data we can use the $ operator: df$sex will give us the sex column. 
With dataframes, we can call specific rows and collumns using the index. For example with df[1,6], will give us the 1st row and the 6th column. We can also select a range of columns or rows using `:` operator. 

```{r}
#Try calling the wt column using the `$`
df$wt
#Try another way to get the wt column
df[,7]
#Try to get the first ten rows of the weight column
df[1:10,7]
```


2. There are some base functions that allow us to visualize our data:
`hist()` will provide a historgram of a variable.
`plot()` can allow you to compare two variables, x and y, in many different types of plots. 

```{r}
#Try plotting wt as a histogram
hist(df$wt)
#try plotting ht as the x variable and wt as the y variable
plot(x=df$ht, y=df$wt)
#Now lets use the help function ? to view the plot function
?plot
#Lets try changing the type argument
plot(x=df$ht, y=df$wt, type = "p")
```

#Introduction to data tidying and tidyverse




