---
layout: page
title: "Introduction to Machine Learning in R"
visible: false
tags:
  # languages
  - r
  # levels
  - intermediate
---

 - **Authors**: Raul Samayoa ([@ralle123](https://github.com/ralle123))
 - **Lesson topic**: Machine learning
 - **Lesson content URL**: <https://github.com/UofTCoders/studyGroup/tree/gh-pages/lessons/r/ml-intro>

Slides are available [at this link](../intro-machine-learning.pdf)

## Demo examples:

### Linear regression

```{r}
#install.packages("lattice")
library(lattice)
alligator <- data.frame(
  lnLength = c(3.87, 3.61, 4.33, 3.43, 3.81, 3.83, 3.46, 3.76,
    3.50, 3.58, 4.19, 3.78, 3.71, 3.73, 3.78),
  lnWeight = c(4.87, 3.93, 6.46, 3.33, 4.38, 4.70, 3.50, 4.50,
    3.58, 3.64, 5.90, 4.43, 4.38, 4.42, 4.25)
  )

# Plot our information
plot(
    alligator$lnWeight,
    alligator$lnLength,
    col = "blue",
    xlab = "Snout vent length (inches) on log scale",
    ylab = "Weight (pounds) on log scale",
    main = "Alligators in Central Florida"
    )

```

[lm](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html) function
fits a linear model. A typical model has the form response ~ terms

```{r}
myModel <- lm(lnLength ~ lnWeight, data = alligator)
```

Visualize the regression

```{r}
# Plot the chart with lm.
plot(
    alligator$lnWeight,
    alligator$lnLength,
    col = "blue",
    abline(myModel, col = 'red'),
    xlab = "Snout vent length (inches) on log scale",
    ylab = "Weight (pounds) on log scale",
    main = "Alligators in Central Florida"
    )

```

```{r}
# Find weight of a alligator with weight of 5.5.
find_value <- data.frame(lnWeight = 5.5)
result <- predict(myModel, find_value)
result
```

If we trace the line with the value we are looking for we can see that the value
seems correct

```{r}
library(ggplot2)
end <- ggplot(alligator, aes(x = lnWeight, y = lnLength)) + 
    geom_point()
end + geom_smooth(method = 'lm', se = FALSE) + 
    geom_vline(xintercept = 5.5) + 
    geom_hline(yintercept = 4.06312, color = 'red')
```

### k-means

```{r}
library(datasets)
head(iris)
ggplot(iris, aes(Petal.Width, Petal.Length)) + 
    geom_point()
```

After resetting the seed I do a
[kmeans](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html)
just selecting the 3-4 columns of iris, collect 3 iterations, selecting 20
random sets

```{r}
set.seed(20)
irisCluster <- kmeans(iris[, 3:4], 3, nstart = 20)
#Comparing cluster with species
table(irisCluster$cluster, iris$Species)
```

We can graph our result

```{r}
irisCluster$cluster <- as.factor(irisCluster$cluster)
ggplot(iris, aes(Petal.Length, Petal.Width, color = irisCluster$cluster)) + 
    geom_point()
```

Why did we do 3 clusters and not 4? dataset only has 3 species

```{r}
ggplot(iris, aes(Petal.Length, Petal.Width, color = Species)) + 
    geom_point()
```
