---
title: "Linear regression"
       "supervised learning"
---
https://stat.ethz.ch/R-manual/R-devel/library/stats/html/lm.html
```{r}
rm(list = ls())
#install.packages("lattice")
library('lattice')
alligator <- data.frame(
  lnLength <- c(3.87, 3.61, 4.33, 3.43, 3.81, 3.83, 3.46, 3.76,
    3.50, 3.58, 4.19, 3.78, 3.71, 3.73, 3.78),
  lnWeight <- c(4.87, 3.93, 6.46, 3.33, 4.38, 4.70, 3.50, 4.50,
    3.58, 3.64, 5.90, 4.43, 4.38, 4.42, 4.25))

# Plot our information
plot(alligator$lnWeight,alligator$lnLength,col = "blue",xlab = "Snout vent length (inches) on log scale",
  ylab = "Weight (pounds) on log scale",main = "Alligators in Central Florida")

```
lm function fits a linear model. A typical model has the form response ~ terms 
```{r}
myModel <- lm(lnLength ~ lnWeight, data = alligator)
```
Visualize the regression
```{r}
# Plot the chart with lm.
plot(alligator$lnWeight,alligator$lnLength,col = "blue",abline(myModel,col = 'red')
,xlab = "Snout vent length (inches) on log scale",ylab = "Weight (pounds) on log scale",main = "Alligators in Central Florida")

```
```{r}
# Find weight of a alligator with weight of 5.5.
find_value <- data.frame(lnWeight = 5.5)
result <-  predict(myModel,find_value)
print(result)

```
If we trace the line with the value we are looking for we can see that the value seems correct
```{r}

library('ggplot2')
end<-ggplot(alligator,aes(x=lnWeight,y=lnLength)) + geom_point()
end<-end + geom_smooth(method = 'lm',se=FALSE) + geom_vline(xintercept = 5.5) + geom_hline(yintercept = 4.06312,color = 'red')
end
```
---
title: "k-means"
       "unsupervised learning"
---
https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html
```{r}
library(datasets)
head(iris)
first<-ggplot(iris,aes(Petal.Width,Petal.Length))+geom_point()
first
```
After resetting the seed I do a kmeans just selecting the 3-4 columns of iris, collect 3 iterations, selecting 20 random sets
```{r}
set.seed(20)
irisCluster <- kmeans(iris[, 3:4], 3, nstart = 20)
#Comparing cluster with species
table(irisCluster$cluster, iris$Species)
```
We can graph our result
```{r}
irisCluster$cluster <- as.factor(irisCluster$cluster)
ggplot(iris, aes(Petal.Length, Petal.Width, color = irisCluster$cluster)) + geom_point()
```

Why did we do 3 clusters and not 4? dataset only has 3 species
```{r}
ggplot(iris, aes(Petal.Length, Petal.Width, color = Species)) + geom_point()
```
