---
title: "Clustering_student"
author: "Tara"
date: "2/9/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction

Today we will learn two techniques that we can use to explore our data and find patterns in our dataset. Clustering analysis is a type of unsupervised machine learning that we use to find patterns or groupings in the data.

The two methods we are learning are hierarchical agglomerative clustering and k-means clustering. The main difference between these two methods is that with hierarchical clustering, we choose the number of clusters we want a posteriori (or after we conduct the analysis), for k-means clustering, we determine the number of clusters we want a priori (prior to the anlalysis). In addition to the choice of the two clustering methods, there are a few methodological decisions we can make that will affect our results, we will dicuss some of these today. 

##Setting up the data and preprocessing
```{r}
#Load in a dataset that is already in R

#Use the help function to learn about the dataset

#What types of variables do you see? Which ones do you find interesting? we can conduct clustering analysis on a subset of the variables or on the entire dataset because we have numeric (including two binary) variables
```

```{r}
#preprocessing the data
#1 check for missing values

#this data set doesn't have any missing values so we can proceed with our analysis
#Normalization and Scaling, chose either: 
  #Normalize function from library(BBmisc)
  #install.packages("BBmisc")
 
  #scale the data using scale()

```


##Hierarchical Agglomerative Clustering
For hierarchical agglomerative clustering begins with each observation in a distinct (singleton) cluster, and successively merges clusters together until a stopping criterion is satisfied. This approach can be visualized as a dendroagram, a tree-like diagram, showing how pairs of observations are fused together. 

There are three different method choices that will affect the way our tree looks: 
  (1) distance metric. this is the calculation that determines how similar the pairs are to eachother. 
  (2) different algorithms for how to etermine how close pairs of observations are to eachother. This will give us different shapes of dendrogrmas. 

```{r}
#Step 1. Create the distance matrix
  #using Euclidean distance, this is most common for data that has continuous numerical values
  #what other methods are there?

  #we can also look at the maximum distance, manhattan distance, canberra, binari, and minkowski. Let's also try the maximum distsance, create another distance matrix adn change the method agreemeent to "maximum"
```

```{r}
#Step 2. Plot the clustering algorithms
#Run the code for how many clusters does it look like there are in each dendrogram?
#complete method
#average method
#single method
#ward.D2

#which method gives the most balanced?

#try the exercise with the dist_Max matrix
```

```{r}
#Step 3. Choosing the number of clusters Storing the Cluster ID for each observation

#using the call bellow, we are taking our clustering and "cutting" the tree into four clusters. The call will return a vector where each observation is assigned to a cluster.


#we can append this vector to a copy of the original dataframe to interpret the clusters

```

```{r}
#Step. 4 Interpretation
#how many cars are in each cluster?

#let's summarize by the groups of clusters
  #need dplyr package


#compare the mean mpg between clusters


#on your own compare another variable mean between clusters

#visualize the clusters 


```


##Kmeans Clustering
K= the number of centres (the number of clusters that we specify apriori).
K-means aims to divide data into K groups (clusters). This involves finding K centroids. Centroids are invented or existing points that represent the centers of the clusters. In K-means clustering the aim is to indetify K number of centroids and allocate every object to the nearest cluster.

The algorithm for K-means is as follows:

1. Place K points into the object data space. These points represent the initial group of centroids,

2. Assign each data point to the closest centroid, based on Euclidean distance.

3. Recalculate the positions of the K centroids. This is done by taking the mean of all data points assigned to that centroid’s cluster.

4. Repeats steps 2 and 3 until the posistions of the centroids no longer change and the sum of distances of individual units from centroids is as small as possible.

Here is a pipeline for conducting K-Means analysis. 

Step 1: Preprocessing (scaling/normalization)
Step 2: Figure out the optimal K-value
Step 3: Run KMeans algorithm
Step 4: Interpret


###Step 2
```{r}
##Step 2: Figure out the optimal K-value. 
#Different methodological approach: trial and error, choose a priori based on theoretical knowledge, choose based on the hierarchical agglomerative clustering result, or use an numerical approach: elbow plot, gap statistic, silohuette.
#we can use these packages to do all three of these numerical approaches
#install.packages("factoextra")
#install.packages("NbClust")

#elbow method


#silohuette


#gap statistic, 


```

##Step 3: Run KMeans algorithm
```{r}
#set the seed for the analysis, for reproducibility
#run the algorithm to start the clusters
#try to look at what the results are in the kclust object

```

###Step 4: Interpret

```{r}
#save the scaled data as a dataframe
#create a new column as of the clustering labels 
#now you group by the clusters, run stats or visualize 
```


##Heatmaps and other visualization methods (time permitting)

```{r}
#the factoextra package has many helpful functions for computing cluster analyses and visualizing them
#another way to visualize clusters, helpful package 
#note that the dimensions are created using Principal components analysis first
fviz_cluster(kmeans(dfScaled, centers = 2), geom = "text", data = dfScaled)
```

##Heatmaps

##Notes and other helpful resources: 
PAM (Partitioning around mediods) works similar to kmeans but, uses a mediod (median) instead of the mean to create clusters. Kmeans uses the centroid, which is an artificial point, while PAM uses the mediod (a real point). 

Here is a helpful textbook to learn more about the theory of these clustering techniques and applying unsupervised and supervised machine learning techniques in R:

James, Witten, D., Hastie, T., & Tibshirani, R. (2013). An introduction to statistical learning: With applications in R (Vol. 103). Springer. https://doi.org/10.1007/978-1-4614-7138-7

This book is available for free from the UofT library at
https://librarysearch.library.utoronto.ca/permalink/01UTORONTO_INST/fedca1/cdi_askewsholts_vlebooks_9781461471387


