---
title: Data visualization in R with ggplot2
author: Ahmed Hasan and James Santangelo, borrowing heavily from Joel Ostblom
output: pdf_document
---
  
## Lesson preamble
  
> ### Learning Objectives
>
> - Produce scatter plots, histograms, bar plots,
> and boxplots using ggplot.
> - Plot model fits with `geom_smooth`.
> - Set universal plot settings.
> - Understand and apply faceting in ggplot.
> - Customize plots with `theme`.
> 
> ### Setup/Required packages
>
> - `install.packages('dplyr')`
> - `install.packages('ggplot2')`
>


```{r, echo=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```
-----
  
# Plotting with ggplot2

## Introduction

`ggplot2` is a plotting package that makes it simple to create complex plots
from data frames. The name `ggplot2` comes from its inspiration, the book "A
grammar of graphics", and the main goal is to allow coders to express their
desired outcome on a high level instead of telling the computer every detail
about what will happen. For example, you would say "color my data by species"
instead of "go through this data frame and plot any observations of species1 in
blue, any observations of species2 in red, and so on". Thanks to this
functional way of interfacing with data, only minimal changes are required if
the underlying data change. This helps in thinking about the data and creating
publication quality plots with minimal amounts of adjustments and tweaking.

`ggplot2` graphics are built step by step by adding new elements, or layers.
Adding layers in this fashion allows for extensive flexibility and
customization of plots. 

```{r}
library(ggplot2)
```

## The `diamonds` dataset

We will be working with the `diamonds` dataset in this lesson. This is a large
(~54,000 x 10) dataset containing measurements and prices of various diamonds.
If you have an older computer and would like to just work with a subset of the
data, you can create that subsetted data frame as follows:
  
```{r}
d <- head(diamonds, n = 1000)
```

This will create a new data frame `d` containing just the first 1000 rows
of the dataset.

## Plotting with `ggplot2` - the geoms

### Making our first scatter plot

To build a ggplot, we need to:
  
1. Use the `ggplot()` function and bind the plot to a specific data frame using
the `data` argument

```{r}
ggplot(data = diamonds)
```

If the arguments are provided in the right order then the
names of the arguments can be omitted (as with `dplyr`)

```{r}
ggplot(diamonds)
```

2. Define aesthetics (`aes`), by selecting the variables to be plotted and the
variables to define the presentation such as plotting size, shape color, etc.

```{r}
ggplot(diamonds, aes(x = carat, y = price))
```

3. Add `geoms` -- geometrical objects as a graphical representation of the data
in the plot (points, lines, bars). `ggplot2` offers many different geoms; we
will use a few common ones today, including:
* `geom_point()` for scatter plots, dot plots, etc.
* `geom_line()` for trend lines, time-series, etc.
* `geom_histogram()` for histograms

To add a geom to the plot, use the `+` operator. Because we have two continuous
variables, let's use `geom_point()` first:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
geom_point()
```

The `+` in the `ggplot2` package is particularly useful because it allows you
to modify existing `ggplot` objects. This means you can easily set up plot
"templates" and conveniently explore different types of plots, so the above
plot can also be generated with code like this:

```{r, first-ggplot-with-plus}
# assign plot to a variable
diamonds_plot <- ggplot(diamonds, aes(x = carat, y = price))

# draw the plot
diamonds_plot + geom_point()
```

Notes:

- Anything you put in the `ggplot()` function can be seen by any geom layers
  that you add (i.e., these are universal plot settings). This includes the x
and y axis you set up in `aes()`.
- You can also specify aesthetics for a given geom independently of the
  aesthetics defined globally in the `ggplot()` function.
- The `+` sign used to add layers must be placed at the end of each line
  containing a layer. If, instead, the `+` sign is added in the line before the
other layer, `ggplot2` will not add the new layer and R will return an error
message.

```{r}
# this will raise an error
ggplot(diamonds, aes(x = carat, y = price)) 
+ geom_point()
```


### Building plots iteratively + scatter plots

Building plots with `ggplot2` is typically an iterative process. We start by
defining the dataset we'll use, lay the axes, and choose a geom:
  
```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point()
```

Then, we start modifying this plot to extract more information from it. For
instance, we can add transparency (`alpha`) to reduce overplotting:
  
  
```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.5)
```

From the eye test, we can see some evidence of clusters in this scatter plot.
It's possible that one of the categorical variables we have in the data could
explain this pattern. Coloring the data points according to a categorical
variable is an easy way to assess if there seems to be correlation. Let's try
this with `cut`.

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.5)
```

While some clustering is visible here, the pattern is not especially clear.
Perhaps something else in our dataset might be a better predictor. Let's try
`clarity` instead:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
geom_point(alpha = 0.5)
```

Turns out clarity appears to be a strong predictor here after all.  Given that
IF1 is the best possible clarity measurement, we see that even 1-2 carat IF1
diamonds fetch equivalent prices to 4-5 carat I1 (lowest clarity) diamonds.

### Histograms

Similarly to scatter plots above, histograms in `ggplot2` can simply be tacked
on as another geom. The difference is that histograms do not take in a y
aesthetic, since that'll be 'count' by definition:

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram()
```

The binning of a given histogram can be modified by either specifying a
binwidth or a fixed number of bins:

```{r}
ggplot(diamonds, aes(x = carat)) +
  geom_histogram(binwidth = 0.05)
```

### Boxplots

Let's go back to the pattern we saw earlier, where `clarity` seemed to be related to `carat`.
We might ask: what is the distribution of `carat` for each value of `clarity`? This is where
a boxplot might be handy:

```{r}
ggplot(diamonds, aes(x = clarity, y = carat)) +
  geom_boxplot()
```

## Model fits with `geom_smooth`

Fitting models to data is a common means of investigating whether certain
relationships are at play in a given dataset. `ggplot2` makes visualizing
these easy by largely consolidating line-fitting into a single geom -
`geom_smooth`.

By default, `geom_smooth` will automatically select what it thinks is the best
model for your dataset. However, rarely will it select a linear fit. To draw a
linear fit (or any other sort of fit, i.e. LOESS, quadratic, etc.) we have to
specify it:

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm')
```

Adding the color aesthetic we had previously will cause lines to be fit to each individual group:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = 'lm')
```

To still fit a line across the entire dataset, specify the color aesthetic _within_ `geom_point`, and not globally.

```{r}
ggplot(diamonds, aes(x = carat, y = price)) +
  geom_point(aes(color = clarity), alpha = 0.5) +
  geom_smooth(method = 'lm')
```

## Intermediate `ggplot` 

### Customizing plot appearance

`ggplot2` also comes preloaded with a host of pre-made themes - try them out:

```{r}
p + theme_bw()
p + theme_classic()
```

There are also ways to create your own custom themes using the `theme()`
function, but that is a bit outside of the scope of this lesson. There are
plenty of excellent resources available online though, some of which are linked
at the end of this lesson! 

### Faceting

Finally, `ggplot` has a special technique called *faceting* that allows the
user to split one plot into multiple subplots based on a variable included in
the dataset. We will use `facet_wrap` to make the same scatter plot as above
(price by carat) but split it by clarity:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ clarity) +
  theme_classic()
```

Alternatively, it may be useful to facet by two categorical variables. This is
where `facet_grid` comes in:

```{r}
ggplot(diamonds, aes(x = carat, y = price, color = clarity)) +
  geom_point(alpha = 0.5) +
  facet_grid(cut ~ clarity) +
  theme_classic()
```

# Wrapping up

We've covered many of the core elements of `ggplot2` today:

* Basic structure of a ggplot call
* The `geom` family
    * `geom_point`
    * `geom_histogram`
    * `geom_boxplot`
* Model fits with `geom_smooth`
* Customizing plots with `theme`
* Faceting with `facet_wrap` and `facet_grid`

## Further resources

* EEB313 - [Introduction to `dplyr` and `ggplot2`][eeb313]
* Data Carpentry's [Data visualization with `ggplot2`][dc-r]
* The [data visualization][data-vis] chapter from Hadley Wickham and Garrett Grolemund's _R for Data Science_
* Hadley Wickham's [_ggplot2_ book][ggplot-book]
* Winston Chang's [_R Graphics Cookbook_][cookbook]

[cookbook]: http://www.cookbook-r.com/Graphs/
[data-vis]: https://r4ds.had.co.nz/data-visualisation.html
[dc-r]: https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html
[eeb313]: https://uoftcoders.github.io/rcourse/lec04-dplyr.html
[ggplot-book]: https://www.amazon.com/dp/0387981403/ref=cm_sw_su_dp?tag=ggplot2-20


