---
title: "PLOTTING IN R: GGPLOT2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# INTRODUCTION TO GGPLOT2
* ##### Author: Lindsay Coome
* ##### Lesson Topic: Plotting in R using ggplot2

## WHAT IS GGPLOT?
ggplot2 is meant to be an implementation of the Grammar of Graphics, hence gg-plot. The basic notion is that there is a grammar to the composition of graphical components in statistical graphics, and by direcly controlling that grammar, you can generate a large set of carefully constructed graphics tailored to your particular needs. Each component is added to the plot as a layer.

## COMPONENTS OF A GGPLOT2 PLOT
Plots convey information through various aspects of their aesthetics. Some aesthetics that plots use are:

* x position
* y position
* size of elements
* shape of elements
* color of elements

The elements in a plot are geometric shapes, like

* points
* lines
* line segments
* bars
* text

Some of these geometries have their own particular aesthetics. For instance:

**points**

* point shape
* point size

**lines**

* line type
* line weight

**bars**

* y minimum
* y maximum
* fill color
* outline color

**text**

* label value

##LET'S GET STARTED!

Here is an example of the base plotting functions in R:
```{r cars}
plot(cars)
hist(cars$speed)
```

[But ggplot2 plots are prettier...](https://www.google.ca/search?q=ggplot2&client=safari&rls=en&source=lnms&tbm=isch&sa=X&ved=0ahUKEwisuv6V6IvSAhVk0oMKHTNkARkQ_AUICCgB&biw=1279&bih=621#tbm=isch&q=ggplot2+examples)

We'll investigate the dataset we're going to use to make some graphs....
```{r}
library(ggplot2)
str(diamonds)
```

Our first graph?
```{r}
ggplot(diamonds, aes(x=cut, y=carat))
```

We've defined the structure of our graph, but we need to add a "layer" to it in order to define what type of graph it is. Let's make a scatterplot:
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + 
  geom_point()
```

You can also save the structure of the graph to a variable and then add layers and other attributes to it. Let's do that:
```{r}
g <- ggplot(diamonds, aes(x=carat, y=price))
g + geom_point()
```

We can add other "layers" of graphs to our first graph. Let's add a trend line to our scatterplot:
```{r}
g + geom_point() + geom_smooth() 
```

We can also "facet" our graphs, which means dividing a plot into subplots based on the values of one or more discrete variables.
```{r}
g + geom_point() + geom_smooth() + facet_grid(cut~.)
```
