---
layout: page
title: "Intermediate topics in R: Strings and factors"
visible: false
tags: 
  # languages
  - r
  # levels
  - intermediate
---

```{r, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Strings, factors, and regular expressions

 - **Authors**: Luke Johnston (@lwjohnst86)
 - **Research field**: Nutritional Sciences
 - **Lesson topic**: Working with character/string data
 - **Lesson video link**: <https://youtu.be/jdMOmg9SMEc>
 
Install these packages:

```{r, eval=FALSE}
install.packages(c("stringr", "forcats"))
```

Base R has some pretty decent string manipulation and searching functionality.
The primary way of working with strings is through 
[regular expressions](http://www.regular-expressions.info/). Regular
expressions, or regex in R, are a group of characters that have special meaning
and function. For instance, the `.` means wildcard and it will search for any
character (letter, number, symbol, etc). Or `$` means end of the string or line.
If you want to see more details on regex in R, run `?regex` to get help on it.

In general, the most commonly used functions (at least that I use and that I've
seen) are `grep`, `grepl`, `sub`, and `gsub` commands. Each of these take regex
characters (`$ ^ * . + ? [] () |` are a few; see `?regex` for more). Let's
begin with the base R functions!

```{r}
fruits <- c("apple", "banana", "pear")

# replace `a` at beginning of string with nothing. `gsub` = global substitution
gsub("^a", "", fruits)
# replace letters `an` or `ar` with `bb`
gsub("(an|ar)", "bb", fruits)
# same as above but only first search and replace. `sub` = substitution
sub("(an|ar)", "bb", fruits)
# search for fruits starting with a. Returns the number position of the search
# result.
grep("^a", fruits)
# same as above, but returns the actual value rather than the number
grep("^a", fruits, value = TRUE)
# same as above, but returns a vector of logical values
grepl("^a", fruits)
```

While base R is pretty powerful, it isn't always easy to work with and there are
some major bits of functionality missing. That's where the `stringr` package
(which is a subset of common functions from the `stringi` package) comes into
play. `stringr` greatly expands on string manipulation in R. Let's go through
some.

```{r}
library(stringr)

# month.name is a default variable in R. I'm putting it in a new variable to
# make it shorter to type out.
mth <- month.name
mth

# Almost every stringr function begins with `str_`.
# Convert cases (lower case, UPPER CASE)
str_to_lower(mth)
str_to_upper(mth)

# Find out the number of characters in each string of the vector
str_length(mth)
mth

# Combine (concatenate) the strings together. Similar to `paste`
str_c(mth, collapse = ", ")

# Take a subset of the months based on a regex.
# for months starting with `J`
str_subset(mth, "^J")
# for months ending in y
str_subset(mth, "y$")
# for months that contain either s, n, or d (only lower case)
str_subset(mth, "[snd]")
# same as above, but all letters since months have been converted to lower case
str_subset(str_to_lower(mth), "[snd]")

# Because stringr is part of the tidyverse packages, you can also use the pipe
# `%>%` command with the commands, to make it easier to read.
mth %>%
    str_to_lower() %>%
    str_subset("[snd]")

# Using the fruit variable in the stringr package... different from the `fruits`
# we created above!
# str_detect converts to logical, for use in filtering, subsetting, etc.
# searches for all fruit beginning with `be`
str_detect(fruit, "^be")

# Find out the proportion (mean of values that are either 0 or 1) of fruit with
# at least one vowel at the end of the string.
str_detect(fruit, "[aeiou]$") %>%
    mean()

# search for any fruit that does not end in an `a`
str_detect(fruit, "^.*[^a]$")
# this is the same as above, but subsets instead
str_subset(fruit, "[^a]$")

# similar to sub
# search any month that starts with J and ends with e and replace the middle
# letters with `HIII`.
str_replace(mth, "^(J).*(e)$", "\\1HIII\\2")

# vector of sentences
head(sentences)

# you can split strings up by a specific character. `str_split` creates a list,
# so I use `[[1]]` or `[1]` to show only the first one.
str_split(sentences, " ")[[1]] # this outputs a vector
str_split(sentences, " ")[1] # this outputs a list
# or split by the letter `e`
str_split(sentences, "e")[[1]]
# if you want to split by the word, use the `boundary` function.
str_split(sentences, boundary("word"))[[1]]
# or by a sentence
str_split(sentences, boundary("sentence"))[[1]]

# include in a pipe
str_split(sentences, boundary("sentence"))[[1]] %>%
    # convert to Title Case
    str_to_title()
```

String manipulation is great, but sometimes you need to convert to factor
variables for use in statistical models or when plotting (for instance when
using `ggplot2`). The `forcats` package makes it easier to work with factors.
Base R really only has one function for working with factors... the `factor`
command.

```{r}
library(forcats)

str(mth)
# This is base function. Compare this...
as.factor(mth)
# with this...
as_factor(mth)
# base R as.factor orders alphabetically, while forcats as_factor orders in
# appearance.

mth2 <- c(mth[10:12], mth[6:12], mth)
mth2

# you can also order by frequency:
fct_infreq(mth2)
# this more explicitly orders by appearance.
fct_inorder(mth2)

# this counts the number of levels within the factors. It is a simple summary
fct_count(mth2, sort = TRUE)

# if you want to convert levels that contains only a few (in this case, less the
# 10% of the total proportion), they will be converted to `Other`. This is a
# nice, quick function to group together uncommon values.
fct_lump(mth2, prop = 0.1)
# or any level with only an n=1 in the category
fct_lump(mth2, n = 1)
# or if you want them renamed to `None`
fct_lump(mth2, prop = 0.1, other_level = 'None')

# this is also a really nice function to rename levels within a factor.
fct_recode(mth2,
           Jan = "January",
           Feb = "February",
           "Dec ember" = "December"
           )

```


# Resources

Very good book on R and data (<https://r4ds.had.co.nz>). Also the
[tidyverse](https://tidyverse.org) resource is good.

Factors (`forcats`):

- <https://forcats.tidyverse.org/>
- <https://r4ds.had.co.nz/factors.html>

Strings (`stringr`):

- <https://stringr.tidyverse.org/>
- <https://r4ds.had.co.nz/strings.html>
- For more complex string work, check out the
[stringi package](https://cran.r-project.org/web/packages/stringi/index.html)

