---
title: Creating websites with RStudio and blogdown
author: Vicki M. Zhang, Ahmed Hasan
date: '2021-12-01'
---

-   **Authors**: Vicki M. Zhang, Ahmed Hasan
-   **Lesson Topic**: Creating websites with RStudio and blogdown

```{r include = FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = FALSE, collapse = TRUE, comment = NA, tidy = FALSE)
```

# Introduction

## Why create a static site?

With the multitude of website-building platforms touting a "website in less than
5 minutes", there are plenty of ways to put up a personal site to showcase your
research. However, there are several reasons to switch from something like Wix,
WordPress or SquareSpace to your own static site.

Firstly, **no ads**. Usually, website generations require you to upgrade your
subscription to remove ads from your site. Secondly, **you can fix your own bugs**. 
In contrast, the bugs that might plague your generated site - such as misbehaving
widgets or inconsistent behaviours across platforms - might be
subject to the developers to fix. But thirdly, and I think most importantly,
**learning a new thing is fun**. Making your own static site using RStudio and
`blogdown` is a very rewarding challenge.

## What is blogdown?

The package that we are using is `blogdown`, developed by [Yihui
Xie](https://yihui.org/en/), [Amber Thomas](https://amber.rbind.io/), [Alison
Presmanes Hill](https://alison.rbind.io/). It allows users to generate a static
website using R markdown files. Note that today, we won't have time to push
everything to GitHub, which serves our website on the web. We will only be
creating our site locally - please refer to the [Collaborating with
GitHub](https://uoftcoders.github.io/studyGroup/lessons/git/collaboration/lesson-AH/)
workshop for more on how to push your website to GitHub!

## What is Hugo?

Hugo is a open-source static site generator. It comes with multiple themes, and
it is fully customizable via **widgets**, which are content blocks on a webpage.
The theme that we are using today is the [Hugo
Lithium](https://github.com/yihui/hugo-lithium) theme, which is a relatively simple
but flexible template to create a first website with. More themes can be found on
the [Hugo themes gallery](https://themes.gohugo.io/). 

# Installing 

If you don't already have blogdown installed, make sure to do so:

```{r}
install.packages('blogdown')
```

You also need to create a project in R that will store all of your website's files. To do this, click:

Click File > New Project > New Directory > New Project.

Then, locate the folder and name the folder that will contain your files (make sure it's empty!).

Click Create Project.

```{r}
library(blogdown)
new_site(theme = "yihui/hugo-lithium") # an optional dir arg can be provided too
```


# Modifying your site

## What do you want your site to look like?

Let's assume we are making a simple personal academic website. From this follows a structure
when we consider the things we might want on there: for instance, a CV page, a blog posts page,
and perhaps also a contact page. 

Widgets are sections on your page that contain information, and different widgets have different layouts and can might be useful for different things. You can have separated pages with only a couple widgets or one page with a lot of widgets. 

### The `config.toml` file

No matter how we want our website to look, modifying any blogdown theme starts
with the `config.toml` file. Every blogdown theme will have one of these (note
that in rare instances, the file may be called `config.yaml` instead).

In the Hugo Lithium theme, this file will be found in the root of the project (it may be
in a `content` folder or something similar in other themes) and should look like a series
of key-value pairs - something like this:

```
title = "A Hugo website"
theme = "hugo-lithium"
googleAnalytics = ""
disqusShortname = ""
```

Don't worry if some of the keys/values there don't make a lot of sense right now -
turns out a lot of these are set to sensible defaults and likely won't need to
be modified. For now, let's modify the title and input our socials in the designated
parts of the file:

```
title = "My Hugo website"
theme = "hugo-lithium"
googleAnalytics = ""
disqusShortname = ""
ignoreFiles = ["\\.Rmd$", "\\.Rmarkdown$", "_cache$", "\\.knit\\.md$", "\\.utf8\\.md$"]

[permalinks]
    post = "/:year/:month/:day/:slug/"

[[menu.main]]
    name = "About"
    url = "/about/"
[[menu.main]]
    name = "GitHub"
    url = "https://github.com/UofTCoders"
[[menu.main]]
    name = "Twitter"
    url = "https://twitter.com/UofTCoders"
```

We'll be revisiting this file over the course of the lesson as we modify our website further.
For now, however, let's save the `.toml` file and then get a live look at these changes.
For this purpose, blogdown includes a function called `serve_site` that'll render a local
instance of your website right here in RStudio. Let's give it a go:

```{r}
blogdown::serve_site()
```

This local instance of your website will show up in the Viewer pane in the bottom right of your RStudio
instance. Note that **this is live** - that means if you make any changes to your website, they will
be immediately reflected in the viewer! This also means `serve_site()` only has to be run once, which
simplifies things a good bit. 

## Working with pages

### Updating an existing page

We've now updated the GitHub and Twitter links seen on the top right of our website,
but if we click on About, we see a brief blurb about the theme itself. Let's change that up!
The About page (and all other pages for our websites) is rendered from the `content` folder -
in this case, specifically in a file called `content/about.md`. Let's open that up and modify
the text to something like:

```
This is my new About page. Interests of mine include R and blogdown.
```

After the file is saved, the About page should be updated live in the Viewer
pane. Note that Markdown styling is fully supported here and blogdown will
faithfully render text however it's been styled:

```
This is my new About page. Interests of mine include **R** and **blogdown**. They're _really_ neat.
```

### Creating a new page

Of course, we might want to create pages beyond just the defaults that a theme provides us. Let's
say we want to add a page containing a CV summary. 

First off, we revisit `config.toml`. Remember the following section?

```
[[menu.main]]
    name = "About"
    url = "/about/"
[[menu.main]]
    name = "GitHub"
    url = "https://github.com/UofTCoders"
[[menu.main]]
    name = "Twitter"
    url = "https://twitter.com/UofTCoders"
```

In the Lithium theme, this is where the page structure of our website has been
defined. You might notice that the About page's url is simply `about`. It follows
that if we want to create a new page, say one called 'CV':

1. This part of the config file will need to be modified, and
2. We'll need to make an `.md` file, much like `about.md`, containing the contents of the file.

Fortunately, blogdown simplifies the second step in this process for us with a
function called `new_content`. Let's try it out:

```{r}
new_content('content/cv.md')
```

This should immediately open a new Markdown file in RStudio for us to work with.

Before we add content, you might notice that the top of the file looks something like this:

```
---
title: ''
date: ''
---
```

This is referred to as a **YAML header** and is where we can store some metadata about the file
(or in this case, the page). Let's update it:

```
---
title: 'Curriculum Vitae'
date: '2021-12-01'
author: 'me!'
---
```

We might notice that despite adding a date and author field, neither actually
shows in the Viewer pane! This is because the parts of the YAML header that show
on the page are governed by the theme. Modifying themes is beyond the scope of
this lesson, but definitely something worth learning if you'd like to take your
blogdown journey further and customize your site in detail.

In either case, let's add some content underneath the header:

```
## Research Publications

Me, My cat. (2021) The efficacy of giving cats pats on the head. _Nature_. 
Me, My cat. (2020) Why treats for your cat are always a great idea. _Science_.

## Other projects in progress

Me. (2021) Earl Grey tea provides significant boost to mood. _Cell_.
```

The Viewer pane should continue to update live as soon as we save the file. As
great as this page now looks, we still do need to update the header to add a
link! Here's where things get a little theme-dependent - different themes may
have different approaches to `config.toml` files, and may require a bit of
detective work to modify correctly. Usually, understanding how the existing
template pages are organized is usually enough to understand how to add our own
pages as needed (or even potentially remove some that are included in the
theme).

Let's head back to `config.toml` and add a link to the CV page:

```
[[menu.main]]
    name = "About"
    url = "/about/"
[[menu.main]]
    name = "GitHub"
    url = "https://github.com/UofTCoders"
[[menu.main]]
    name = "Twitter"
    url = "https://twitter.com/rstudio"
[[menu.main]]
    name = "CV"
    url = "/cv/"
```

Here, we have to make sure the url field specifically points to our Markdown file,
which was named `cv.md` (see how About is formatted at the top, for instance).
Once `config.toml` is saved, there should now be a CV tab in the Viewer pane! 

## Making new blog posts

To round out the lesson, we'll have a look at how to make blog posts for your website.
Most blogdown themes will have blogging functionality built in (it's in the name after all)
which means that making new blog posts can be done entirely through a blogdown function
once more.

We make a new post by running:

```{r}
blogdown::new_post(title = "My first post", 
                   ext = '.Rmd', 
                   subdir = "post")
```


This creates a new .Rmd file in a new folder, all in `content/post`, and
automatically navigates to it. Then, in your directory, you will be able to see
the folder with the **slug**, or the unique part of the URL that identifies your
post. In your folder, all of the files needed for this blog post will be
bundled, including the main .Rmd post, as well as any images.

As before, the Viewer should immediately update with our first post, and RStudio
will be primed for editing it. Something that should immediately stick out is the
increased number of fields in the YAML header for a blog post:

```
---
title: My first post
author: ''
date: '2021-12-01'
slug: my-first-post
categories: []
tags: []
---
```

Categories and tags are user-defined and can be used to sort blog posts by topic
and more. For now, however, we'll leave those blank and just modify the author
field at most before we get writing.

Let's add some content. Note that **this is a legitimate R Markdown file**,
and that means that features like code chunks and inline output are all supported!

Let's try that out and see for ourselves:

````
---
title: My first post
author: 'Me'
date: '2021-12-01'
slug: my-first-post
categories: []
tags: []
---

This is my first blog post in blogdown. It essentially allows me to
write an R Markdown file and have it rendered on a website.

Here's some _styling_ using Markdown syntax.

Here's a code chunk:

```{r}`r ''`
2 + 2
```

And a plot:

```{r}`r ''`
plot(iris$Sepal.Length, iris$Sepal.Width)
```

````

Once this file is saved, the website will render the full contents of the Rmd
file and _also run each code chunk_, placing the output of each just below
the corresponding lines of code. That includes plots! This right here is
an especially powerful feature of blogdown - it's very amenable to writing
programming tutorials and the like because of inline output rendering.

If rendering this on a machine that has Python installed, Python code can
be run here as well:

````
```{python}
l = [1, 2, 3]
for i in l:
    print(i, i * 2)
```
````

# Conclusion

In today's lesson, we've covered:

- Creating a blogdown site using a theme
- Understanding the role of the `config.toml` file
- Creating a new page and modifying existing pages
- Creating a new blog post

One important thing we did _not_ cover is how to push your website to a hosting
service (e.g. GitHub, Netlify). This is a bit more involved - but not
necessarily difficult - and a full guide on how to do so can be found in the
[official blogdown guide](https://bookdown.org/yihui/blogdown/deployment.html)
under 'Deployment'.

## Further reading

-   [What is hard about blogdown?](https://community.rstudio.com/t/what-is-hard-about-blogdown/8108)
-   [Up and Running with blogdown](https://alison.rbind.io/post/2017-06-12-up-and-running-with-blogdown/)
-   [Creating a website with the academic theme in blogdown](https://www.kevinzolea.com/post/blogdown/creating-a-website-with-the-academic-theme-in-blogdown/)
-   [Overwhelmed by Hugo academic theme: a beginner's guide](https://andreaczhang.rbind.io/post/my-1st-blogpost/)
-   [Fun blogdown in R to design a personal website](https://annielyu.com/2020/01/12/blogdown-website/)
