---
title: "Reproducible Science with RStudio"
author: "Ahmed Hasan"
output:
  pdf_document: default
  html_document: default
always_allow_html: true
---

# Introduction

This is an introduction to R and RStudio with an emphasis on project management
and the various features RStudio has to make reproducible science easier. We'll
be working through a simple project workflow using the various project management
features RStudio provides, and pick up some basic R along the way. Although no
prior experience is required, it may be helpful to read across the material from
the UofT Coders [Intro to R][intro-r] lesson.

Note: material for an extended version of this lesson, delivered in the Spring
2020 semester, can be found [here][lesson-old]. The version in this file is
truncated and intended to be delivered in 40-50 minutes online, thus
de-emphasizing the code-along sections.

## What does it mean to work reproducibly in RStudio?

Reproducible computational science endeavours to make both the data used in
analyses as well as the code used to process this data available and reusable.
However, ensuring reproducibility is difficult unless certain principles are
adhered to early on; projects may branch off into various unplanned subtasks,
certain scripts may undergo several major changes, leading to cluttered file
directories, and files/scripts may not be named in a consistent manner, making
retracing steps difficult. While these are important considerations when
working with other scientists, it's worth noting that possibly your most
frequent collaborator may be future you, going back to your older work to
figure out how you proceeded with an analysis. It's therefore important to make
sure projects are well managed and documented, and RStudio offers a suite of
features to help facilitate this.

## What is the difference between R and RStudio?

**R** is a statistical programming language, designed primarily with data
analysis and visualization in mind, while **RStudio** is an interface for
working in R. Although it is possible to use R independently of RStudio,
RStudio is widely considered the default platform for working in R due to
its powerful features and ongoing community support. 

## An overview of RStudio

At its core, RStudio consists of four panes (one of which is often hidden upon
startup):

1. Source (text editor)
2. Console (an R interpreter)
3. Environment and history
4. Miscellaneous (files, plots, packages, help docs, etc.)

Most of the time, working in RStudio involves working in the Source pane. This
pane can be opened up by opening either a new or an existing file, and consists
of a text editor. Code in the source pane can be 'sent' to the console to be
run, allowing a user to stay in Source for the most part while they work, but
the Source pane can also be used to edit other text files; this very text is
being written in the Source panel of RStudio! The console is the real engine 
which carries out the commands. All your code is *"run"* through the console
but a copy of the code is kept and organized in the script. 

# R Projects

In this lesson, we'll be making extensive use of yet another feature of RStudio:
**RStudio Projects**. As the name implies, these are designed to keep all files
and scripts associated with a given project, in addition to maintaining
project-specific command history and other useful things.

## Creating a project

Projects can be created using the dropdown in the top right of RStudio. This
menu should currently say `Project: (None)`, but will otherwise reflect the name
of whatever project is open at a given time. Clicking on the dropdown and
selecting `New Project` will open a dialog with three options:

- New Directory
- Existing Directory
- Version Control

We'll be selecting `New Directory` for now. This will create a new folder on your
device that will act as the headquarters for the project. RStudio will ask us to name
the project, and also offer to enable version control with Git for the project.
Although we won't be using Git much in this lesson given time constraints, we
will still check off this checkbox, as it enables a few extra features that are
helpful to know about.

Once the project has been created, RStudio will in fact appear to 'restart' --
the project will now open in a new instance of RStudio. RStudio creates a new 
folder where you specified and also quietly performs a few other operations in the
background:

1. In the newly made project directory, RStudio creates an `.Rproj` file, which
contains various project settings. This can be seen in the Files pane in the
bottom right.
  - This file is also used by certain packages to detect where the project root
  is in the file tree.
2. RStudio also creates a hidden directory in the same location called
`.Rproj.user` containing various project-related temporary files.
3. Since we enabled Git earlier, RStudio should also initialize this folder as a
Git repository and create a `.gitignore` file. If you aren't familiar with Git, don't
worry about this step -- what's essentially happening is that RStudio has primed this folder
for use with Git.
  - There should also be a new 'Git' tab added to the Environment & History pane. 
  
## Setting up our directory structure

One of the most important steps when getting started with a project is deciding upon
and setting up a coherent, understandable directory structure. It's very easy to make
a series of 'temporary' directories that eventually lead to all manner of important
files being scattered every which way -- we've all been there! To avoid this, let's start
with a simple directory structure. 

In the Files pane, use the `New Folder` button to create three directories:
`data`, `analysis`, and `plots`. This directory structure may differ from how
you prefer to do things; some scientists like to have `plots` as a subdirectory
of `analysis`, while others split `data` into `data-raw` and `data-clean`. In
your own work, be sure to experiment with what works for you, but above all: try
to be consistent!

The idea behind this structure is:

- `data` will contain data, as the name implies; however, an important thing to
keep in mind is that every file in this folder should be considered
**read-only**. If data is modified in any way, say after you've filtered out a
few columns in R, these modified datasets should be saved as new files within
this folder.
- `analysis` will contain scripts, notes to oneself (such as digital 'lab
notebooks') and so on. We'll primarily be working from within this folder.
- `plots` will contain any and all data visualizations.

(Again, this is _just an example structure_; you don't have to adhere to this in your own work!)

# Getting started with a sample project

## R Scripts and R Notebooks

Although RStudio is primarily designed to work with R code and scripts, it
also supports a variety of other file formats. If we navigate to the `File`
menu and look at the options under `New`, we can see file formats ranging
from R Notebooks to Markdown files and Python scripts. Any of these file formats
can be used and edited within RStudio (and more importantly for our purposes,
within an R Project).

You may notice that many of these seem to be variants of R scripts, such as
R Markdown, R Notebooks, R Documentation, and more. Of these, perhaps the two
to make note of are R Scripts and R Notebooks.

An **R Script** is simply a plain text file with R code in it and nothing but.
R scripts are very lightweight and perfectly usable for all sorts of R-related
purposes.  They can also be easily run from the command line (e.g. run in their
entirety outside of RStudio) although this is a more advanced technique. 

An **R Notebook**, on the other hand, is a special RStudio-only format that
allows users to easily document and organize code as they go along. R Notebooks
are incredibly powerful and highly recommended for any sorts of analyses in
RStudio, since they make it very easy to ensure that code is well annotated and
presented in a logical manner.

## Using R Notebooks

How do R Notebooks do this? At their core, all R Notebooks are actually
Markdown files. Markdown is a means of styling plain text files using simple
text characters. This lesson material you are reading has actually been written
in Markdown! Since R Notebooks are Markdown files, anything typed into an R
Notebook will be interpreted as Markdown-styled text by RStudio instead of code
(which would be the case with an R Script). 

So that's all well and good, but where does the code come in? Well, R Notebooks
feature the ability to intersperse _code chunks_, which are blocks within
these files that RStudio knows to interpret as code. We can insert a code chunk
in an R Notebook using either Shift + Option/Alt + I or the `Insert` dialog at
the top of the screen.

Let's add in a code chunk and write our first bit of R code -- some simple math:

```{r}
2 + 2
```
We can run the contents of a code chunk by either clicking on the `Run` button
at the top of the screen or by pressing Ctrl/Cmd + Shift + Enter. RStudio will
then execute the code in the console, but will *also* show the output directly
underneath the code chunk. Since each code chunk's output is right next to the
code itself, this means we can work entirely within the source panel if we so
choose! 

(Side note: the **R Markdown** format is effectively identical to R Notebooks,
save for the fact that chunks of code in R Markdown files are run all at once,
while R Notebooks allow for line-by-line running of code. This is why R Notebooks
still have the `Rmd` extension.)

## A sample workflow

R features various object types to hold different kinds of information, such as
numbers and text. We won't go in detail about these object types -- if you're
interested in digging deeper into these topics, we recommend checking out
our [Intro R][intro-r] material. 

For this demo, we'll be working with a dataset on air quality in New York City.
This is a dataset already built into R. The specifics of the code don't matter
as much here; the point of this lesson is to better understand RStudio's
features for reproducible science over working through R itself, so we'll gloss
over some of the details for now.

Opening a new R Notebook, the first thing we notice is the header at the top of
the screen. This is referred to as a **YAML header**, and stores information
about the file itself as well as other related metadata related to knitting. We
will not worry too much about this today, but it's worth knowing what it is for
more advanced R Notebook uses.

The first thing to do in any notebook is *import the packages we are using*.
Packages are extensions of R that provide new functionality. Packages only need
to be installed once, and then loaded in every time they are used. They can
almost be thought of as apps on your phone: you install an app once, but need
to open it every time you use it. 

The package we will be working with today is called `here`. `here` provides a
single added feature that makes it very easy to navigate through the folders in
our project.

```{r}
library(here)
```

We will revisit `here` shortly, but are importing it now because it is good
practice to import packages at the start of a notebook or script. Instead,
let's now look at the `airquality` dataset. This is already loaded into R,
and can be viewed by simply typing in `airquality`:

```{r}
airquality
```

In this example workflow, we are interested in whether recorded wind speed
correlates with the temperature on any given day. Let's plot the two variables
against one another:

```{r}
plot(airquality$Wind, airquality$Temp)
```

We see that there seems to roughly be an inverse correlation. Let's add
a regression line using a linear model:

```{r}
abline(lm(airquality$Temp ~ airquality$Wind))
```

Normally, we would of course do much more in depth statistical work -- for
instance, creating a linear model and checking whether the assumptions of
simple linear regression are met, among other things. For the sake of this
lesson, however, we'll stop here.

Let's say this is our final plot and we want to save it for publication.
Let's first save our notebook to our `analysis` folder.

Now, if we want to save this plot to the `plots` folder, we might be in a bit
of trouble, since we'd need R to go back out one folder (from `analysis` to the
project root) and then into `plots`. This is where the `here` package comes in.
`here` lets us write file paths as if we were always in the project root.
Let's save this plot:

```{r}
png('airquality_plot.png')
plot(airquality$Wind, airquality$Temp)
abline(lm(airquality$Temp ~ airquality$Wind))
dev.off()
```

Checking our folders, it seems that this file has been saved into `analysis` and
not `plots`. Changing the above to `plots/airquality_plot.png` raises an error.
How do we fix this with `here`? 

```{r}
png(here('plots/airquality_plot.png'))
plot(airquality$Wind, airquality$Temp)
abline(lm(airquality$Temp ~ airquality$Wind))
dev.off()
```

There we go -- the plot has now been saved to the `plots` folder! It's good
practice to always use `here` when working with any kind of file paths in R,
whether that involves reading in or saving files. 

# Other, more advanced features

That about does it for what we'll be covering in this lesson, but here's a quick rundown
of a few other features/integrations RStudio offers to help with reproducibility.

## Git integration

RStudio projects are fully compatible with Git, and feature various GUI features to help
facilitate Git usage. Git is a version control system that allows you to track changes in
your files as well as preserve file history. Using Git with your projects means that past
versions of files can be retrieved, and changes can be easily documented. Git alone
could take up an entire lesson itself, however, which is why we haven't covered it today,
but if you're interested in learning how to use it feel free to have a look at our
[Intro Git][intro-git] lesson or check out [Happy Git with R][happy-git-r].

## Packrat

Packrat is a package management system for R that allows users to associate
specific versions of packages with projects. This means that a project can
contain its own self contained packages forever, which is especially useful if
these packages eventually get updated in ways that break compatibility (since
Packrat will specifically allow for the maintenance of whichever version of a
given package was used). More can be found [here][packrat].



[happy-git-r]: https://happygitwithr.com/
[intro-git]: https://uoftcoders.github.io/studyGroup/lessons/git/intro/lesson/
[intro-r]: https://uoftcoders.github.io/studyGroup/lessons/r/intro/lesson/
[lesson-old]: https://github.com/UofTCoders/studyGroup/blob/gh-pages/lessons/r/reproducible-rstudio/lesson.Rmd
[packrat]: https://rstudio.github.io/packrat/


