---
title: "Reproducible Science with RStudio"
author: "Frances Wong and Ahmed Hasan"
output:
  pdf_document: default
  html_document: default
always_allow_html: true
---

# Introduction

This is an introduction to R and RStudio with an emphasis on project management
and the various features RStudio has to make reproducible science easier. We'll
be working through a simple project workflow using the various project management
features RStudio provides, and pick up some basic R along the way. Although no
prior experience is required, it may be helpful to read across the material from
the UofT Coders [Intro to R][intro-r] lesson.

## What does it mean to work reproducibly in RStudio?

Reproducible computational science endeavours to make both the data used in
analyses as well as the code used to process this data both available and
reusable. However, ensuring reproducibility is difficult unless certain
principles are adhered to early on; projects may branch off into various
unplanned subtasks, certain scripts may undergo several major changes, leading
to cluttered file directories, and files/scripts may not be named in a
consistent manner, making retracing steps difficult. While these are important
considerations when working with other scientists, it's worth noting that
possibly your most frequent collaborator may be future you, going back to your
older work to figure out how you proceeded with an analysis. It's therefore
important to make sure projects are well managed and documented, and RStudio
offers a suite of features to help facilitate this.

## What is the difference between R and RStudio?

**R** is a statistical programming language, designed primarily with data
analysis and visualization in mind, while **RStudio** is an interface for
working in R. Although it is possible to use R independently of RStudio,
RStudio is widely considered the default platform for working in R due to
its powerful features and ongoing community support. 

## An overview of RStudio

At its core, RStudio consists of four panes (one of which is often hidden upon
startup):

1. Source (text editor)
2. Console (an R interpreter)
3. Environment and history
4. Miscellaneous (files, plots, packages, help docs, etc.)

Most of the time, working in RStudio involves working in the Source pane. This
pane can be opened up by opening either a new or an existing file, and consists
of a text editor. Code in the source pane can be 'sent' to the console to be
run, allowing a user to stay in Source for the most part while they work, but
the Source pane can also be used to edit other text files; this very text is
being written in the Source panel of RStudio! The console is the real engine 
which carries out the commands. All your code is *"run"* through the console
but a copy of the code is kept and organized in the script. 

# R Projects

In this lesson, we'll be making extensive use of yet another feature of RStudio:
**RStudio Projects**. As the name implies, these are designed to keep all files
and scripts associated with a given project, in addition to maintaining
project-specific command history and other useful things.

## Creating a project

Projects can be created using the dropdown in the top right of RStudio. This
menu should currently say `Project: (None)`, but will otherwise reflect the name
of whatever project is open at a given time. Clicking on the dropdown and
selecting `New Project` will open a dialog with three options:

- New Directory
- Existing Directory
- Version Control

We'll be selecting `New Directory` for now. This will create a new folder on your
device that will act as the headquarters for the project. RStudio will ask us to name
the project, and also offer to enable version control with Git for the project.
Although we won't be using Git much in this lesson given time constraints, we
will still check off this checkbox, as it enables a few extra features that are
helpful to know about.

Once the project has been created, RStudio will in fact appear to 'restart' --
the project will now open in a new instance of RStudio. RStudio created a new 
folder where you specified and also quietly performs a few other operations in the
background:

1. In the newly made project directory, RStudio creates an `.Rproj` file, which
contains various project settings. This can be seen in the Files pane in the
bottom right.
  - This file is also used by certain packages to detect where the project root
  is in the file tree.
2. RStudio also creates a hidden directory in the same location called
`.Rproj.user` containing various project-related temporary files.
3. Since we enabled Git earlier, RStudio should also initialize this folder as a
Git repository and create a `.gitignore` file. If you aren't familiar with Git, don't
worry about this step -- what's essentially happening is that RStudio has primed this folder
for use with Git.
  - There should also be a new 'Git' tab added to the Environment & History pane. 
  
## Setting up our directory structure

One of the most important steps when getting started with a project is deciding upon
and setting up a coherent, understandable directory structure. It's very easy to make
a series of 'temporary' directories that eventually lead to all manner of important
files being scattered every which way -- we've all been there! To avoid this, let's start
with a simple directory structure. 

In the Files pane, use the `New Folder` button to create three directories:
`data`, `analysis`, and `plots`. This directory structure may differ from how
you prefer to do things; some scientists like to have `plots` as a subdirectory
of `analysis`, while others split `data` into `data-raw` and `data-clean`. In
your own work, be sure to experiment with what works for you, but above all: try
to be consistent!

The idea behind this structure is:

- `data` will contain data, as the name implies; however, an important thing to
keep in mind is that every file in this folder should be considered
**read-only**. If data is modified in any way, say after you've filtered out a
few columns in R, these modified datasets should be saved as new files within
this folder.
- `analysis` will contain scripts, notes to oneself (such as digital 'lab
notebooks') and so on. We'll primarily be working from within this folder.
- `plots` will contain any and all data visualizations.

(Again, this is _just an example structure_; you don't have to adhere to this in your own work!)

# Getting started with a sample project

R Markdown is great for keeping your code and outputs organized together. This
is done through code chunks. Text outside of the code chunks, such as this, is
interpreted as plain text. Code needs to be within a code chunk, and the
language said code is written in is specified in the header. They can be
inserted with the green button on the top of this panel or with a keyboard
shortcut (Ctrl/Cmd + Alt + i). Plain text can be added with code chunks by
starting a line with a #.

First step of any project: load in your libraries. It's a good habit to keep
them all together at the start of your document for consistency. Think of
libraries as apps for your phone: you need to download them once the first time
you want to use it (in R, install new packages with `install.packages()` for
base R) but you still need to open it (in R, `library()`) every time you want to
use it. You'll only need to install a package once, but you'll need to load it
in with every new session.

```{r loadingPackages}
# Packages may need to be installed with install.packages() the first time 
library(tidyverse)
library(ggplot2)
library(here)
library(knitr)
library(plotly)
```

Now that we have our first code chunk, the whole segment can be executed using
the green play button at the top right corner of each chunk or by pressing Shift
+ Ctrl/Cmd + Enter. To run a single line within the code chunk, either
highlight it before pressing the green play button or place your cursor on the
line and execute with Ctrl/Cmd + Enter.

Let's start with some basics. R at its core can be used as a giant calculator to
which we can feed variables and equations.

```{r}
2+2
```

Values can also be stored into objects. Notice how the object names and quick
information are populated into the 'Environment' tab when created.

```{r}
bigNumber <- 239
smlNumber <- 3

bigNumber/smlNumber
```

We can start to explore the strength of R by working with more than one value at
once. When an object holds a string of values, they are called vectors. There
are multiple types depending on what kind of information they are holding.
Notice how RStudio has a feature called "syntax highlighting" which will
indicate when different data types are recognized. The actual colors depend on
the theme selected in your settings.

```{r}
numericVec <- c(3, 1, 4, 1, 5)
characterVec <- c("is", "it", "pi", "or", "pie") # character strings must be enclosed in quotations or else they will be interpreted as objects in the environment. 
logicalVec <- c(FALSE, FALSE, TRUE, FALSE, F)
```

Note: logical booleans must be written with all caps (FALSE or TRUE) or by a single capital letter (F or T). 

What happens if there are multiple data types within a vector? 

```{r}
mixedVec <- c("UofT", "Coders", 2020)
mixedVec
class(mixedVec)
```

All values will be coersed into a single data type. 

Everything we've been working with so far has been one dimensional, a vector.
Let's get a little more power with two dimensional data structures. Matrices can
only hold one data type.

```{r}
matrix <- matrix(1:25, nrow = 5) # colon asks for values 1 "through" 25

matrix[5, 3] # Row 5, column 3
```

Matrices can be acted on fully or column-wise:

```{r}
matrix[, 2]*2

matrix[, 2] <- matrix[, 2]*2
```

A second complex data structure is the **data frame**. In contrast to matrices, which can 
only hold one type of data, each column of a dataframe does not need to be the same data type. 

```{r}
df <- data.frame(numericVec, characterVec, logicalVec)
str(df)
```

This allows more flexability in its usage. 

Next, let's bring in a fun dataset to work with: 

```{r readingDataset}
nobel <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-05-14/nobel_winners.csv")
```

This is a collection of metadata for almost all Nobel laureates in physics,
chemistry, and physiology or medicine from 1900 to 2016.

This data is brought in as a modified data.frame specific to tidyverse called a
tibble.

Let's first take a look at the data we have:

```{r exploringDataset}
dim(nobel)
colnames(nobel)
summary(nobel) # not the best for handling characters ... 
```

The most comprehensive one that is really useful for data frames, because
they can hold multiple data types, is `str`, or structure:

```{r}
str(nobel)
```

Let's pretend we just finished data entry and we'd like to save our dataset.
We'll keep a copy in our data folder:

```{r saveSnapshot}
# Using full path 
#write_csv(nobel, "../../nobel_complete.csv")

# or alternatively, using here to navigate from the root of the project folder 
#write_csv(nobel, here("data/nobel_complete.csv"))
```

Using the `here()` function brings the path back to the root directory for the
project. This is great to ensure reproducibility because relative paths are
easily confused when moving between scripts.

Another common practice is commenting out the `write_csv()` line after executing
to prevent accidentally writing over your dataset when re-running code in future
analysis. It is completely up to you if you'd like to adapt this into your
practice!

Looking at one column, for example `category`, we can see that there are a few
possible discrete values. We can take a look at the values and distribution:

```{r exploringCategory}
head(nobel$category)
table(nobel$category)
```

Factors are great, especially for graphing, because they turn values into
discrete categorical data. The different categories are called levels. The
`forcats` package is handy when manipulating factors and level, but that's for
another day!

```{r}
head(nobel$category)

nobel$category <- as.factor(nobel$category)
head(nobel$category)
levels(nobel$category)
```

Let's look at how the category of prize won varies by a laureate's date of birth:

```{r startPlotting}
ggplot(data = nobel, aes(x = birth_date, y = category)) + 
  geom_point()
```

Add a color as a dimension:

```{r}
ggplot(data = nobel, aes(x = birth_date, y = category, color = gender)) + 
  geom_point()
```

Let's look at a different comparison. How does the prize year relate to the
awardee's birth date? We can use the same point plot but add a layer with a fit
on top.

```{r}
ggplot(data = nobel, aes(x = prize_year, y = birth_date, color = category)) +
  geom_point() + geom_smooth()
```

There's too much going on in this plot; let's just look at the data from one
country. Most data in the dataset are from the United States, so let's focus on
that. We'll also save the filtered dataset in case we need to reference it in
the future.

```{r}
nobel_US <- filter(nobel, birth_country == "United States of America")

#write_csv(nobel_US, here("data/nobel_USonly.csv"))

ggplot(data = nobel_US, aes(x = prize_year, y = birth_date, color = category)) +
  geom_point() + geom_smooth()
```

Now we can start to see some trends! 

Let's compare the distribution between the five largest countries in North
America by population.

```{r}
nobel_NA <- filter(nobel, birth_country %in% c("United States of America", "Mexico", "Canada","Guatemala", "Cuba"))

ggplot(data = nobel_NA, aes(x = prize_year, y = birth_date, color = birth_country, shape = category)) +
  geom_point() + 
  theme_classic()
```

The last plot generated can be saved using `ggsave`. 

```{r}
ggsave(here("data/nobel_USonly.png"), width = 6, height = 4)
```

This complex plot can benefit from being intereactive. `ggplot` and `plotly` makes this very 
simple: save the graph into an object and bring it into plotly! Voila! 

```{r}
plot <- ggplot(data = nobel_NA, aes(x = prize_year, y = birth_date, color = birth_country, shape = category)) +
  geom_point() + 
  theme_classic()

ggplotly(plot)
```

Lastly, let's knit this document to have a static snapshot of what we did. The
file (.Rmd) needs to be saved before getting started with the knitting process.
The code also needs to be able to execute 100% perfectly without user input;
errors will break the knitting process. Start knitting by clicking the cleverly
labeld yarn ball at the top of the script console. You'll notice the console
pane where the code was running switched to the R Markdown tab. Keep your eye on
this: any errors and their associated line numbers will be displayed here.

Note: Some devices require additional set up for compiling pdf documents. This may involve 
downloading LaTeX outside of R. 

[intro-r]: https://uoftcoders.github.io/studyGroup/lessons/r/intro/lesson/





